Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems\\ \thanks{Identify applicable funding agency here. If none, delete this.}

### 2024-05-22 abstract.tex
Added framework

@@

Legal cases involving disparate impact often refer to the ``80\% Rule'', advocated by the US Equal Employment Opportunity Commission \cite{eeoc1979}, where it requires the selection rate of a minority group is to be no less than 80\% of that of a majority group. Formally, it is:

@@

\subsection{Demographic Parity}
Demographic parity, also known as statistical parity, is similar to disparate impact but, instead of ratio, difference is taken. It is named so to suggest that each demographic group(such as race, gender, or age) should have equal representation or opportunity. Formally, it is:

@@

\subsection{Conditional Statistical Parity}
Conditional statistical parity is similar to demographic parity, but, in addition to protected attributes, it further takes into account some ``legitimate'' attributes that are \textit{legitimately} related to the case. For example, a legitimate attribute when considering future recidivism could be the number of prior crimes committed. Formally, it is:

@@

\subsection{Overall Accuracy Equality}
Accuracy eqaulity is similar to demographic parity, but instead of the case of $\hat{Y} = 1$, it considers the case of $Y = \hat{Y}$; that is, the case where the prediction is accurate. Formally, it is:

@@

\subsection{Mean Difference}
Mean difference considers the case of non-binary outcome, such as the score of COMPAS. It takes the expected value of the prediction. Formally, it is:

@@

\subsection{Equalized Odds}
Equalized odds is similar to demogrphic parity, but it further takes into account the grouth truth. It consider the cases of true positive and false positive. It solves the downsides of a fully accurate classifier might be deemed unfair by measures that don't take ground truth into consideration. For example, consider a group $A$ that is predicated to recidivate and they do in fact always recidivate and a group $B$ that is predicated to recidivate but end up never recidivating. A fully accurate classifier will always predict $A$ to recidivate while $B$ to never recidivate, and this will violate demogrphic parity. By taking ground truth into account, equalized odds avoids these pitfalls. Formally, it is:

@@

\subsection{Equal Opportunity}
Equal opportunity is a relaxation of equalized odds by only considering the true positive case. It is named so because when outcome being positive is beneficial to the individual, as it is an opportunity, such as school acceptance, the fairness of the true positive case is much more important. Formally, it is:

@@

\subsection{Predictive Equality}
Predictive equality is also a relaxation of equalized odds by only considering the false positive case. It was formulated \cite{corbett2017algorithmic} when considering outcome being positive is detrimental to the individual, such as future recidivism. Thus, the fairness of the false positive case is much more important. Formally, it is:

@@

\begin{table*}[h]
    \centering
    % \resizebox{\textwidth}{!}{
    % \resizebox{\columnwidth}{!}{
    \begin{tabular}{|l|c|} % 'l' specifies left alignment for the first column
        \hline
        \textbf{Column 1} & \textbf{Column 2} \\
        \hline
disparate impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
\hline
demographic parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
\hline
conditional statistical parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
\hline
accuracy eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
\hline
mean difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
\hline
\multirow{2}{*}{equalized odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\cline{2-2}
                                & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
\hline
equal opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
\hline
predictive equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\hline
\multirow{2}{*}{conditional use accuracy equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
\cline{2-2}
                                 & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
\hline
predictive parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
\hline
equal calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
\hline
    \end{tabular}
    % }
    \caption{Example Table}
    \label{tab:example}
\end{table*}

>>>>

\newcommand{\db}{\mathcal{D}}
\newcommand{\model}{\mathcal{M}}
\newcommand{\priv}{P}
\newcommand{\pos}{\hat{P}}
\newcommand{\tru}{T}
\newcommand{\leg}{L}
\newcommand{\sco}{\hat{S}}

@@

Legal cases involving disparate impact often refer to the ``80\% Rule'', advocated by the US Equal Employment Opportunity Commission \cite{eeoc1979}, where it requires the selection rate of a minority group is to be no less than 80\% of that of a majority group. Formally \cite{feldman2015certifying}, it is:

@@

\subsection{Demographic Parity} %TODO: why cite this?
Demographic parity \cite{calders2010three}, also known as statistical parity, is similar to disparate impact but, instead of ratio, difference is taken. It is named so to suggest that each demographic group(such as race, gender, or age) should have equal representation or opportunity. Formally, it is:

@@

\subsection{Conditional Statistical Parity}
Conditional statistical parity \cite{corbett2017algorithmic} is similar to demographic parity, but, in addition to protected attributes, it further takes into account some ``legitimate'' attributes that are \textit{legitimately} related to the case. For example, a legitimate attribute when considering future recidivism could be the number of prior crimes committed. Formally, it is:

@@

\subsection{Overall Accuracy Equality}
Overall accuracy eqaulity \cite{berk2021fairness} is similar to demographic parity, but instead of the case of $\hat{Y} = 1$, it considers the case of $Y = \hat{Y}$; that is, the case where the prediction is accurate. Formally, it is:

@@

\subsection{Mean Difference}
Mean difference \cite{vzliobaite2017measuring} considers the case of non-binary outcome, such as the score of COMPAS. It takes the expected value of the prediction. Formally, it is:

@@

\subsection{Equalized Odds}
Equalized odds \cite{hardt2016equality} is similar to demogrphic parity, but it further takes into account the grouth truth. It consider the cases of true positive and false positive. It solves the downsides of a fully accurate classifier might be deemed unfair by measures that don't take ground truth into consideration. For example, consider a group $A$ that is predicated to recidivate and they do in fact always recidivate and a group $B$ that is predicated to recidivate but end up never recidivating. A fully accurate classifier will always predict $A$ to recidivate while $B$ to never recidivate, and this will violate demogrphic parity. By taking ground truth into account, equalized odds avoids these pitfalls. Formally, it is:

@@

\subsection{Equal Opportunity}
Equal opportunity \cite{hardt2016equality} is a relaxation of equalized odds by only considering the true positive case. It is named so because when outcome being positive is beneficial to the individual, as it is an opportunity, such as school acceptance, the fairness of the true positive case is much more important. Formally, it is:

@@

\subsection{Predictive Equality}
Predictive equality \cite{corbett2017algorithmic} is also a relaxation of equalized odds by only considering the false positive case. It was formulated by considering positive outcome being detrimental to the individual, such as predicting future recidivism. Thus, only the false positive case is taken. Formally, it is:

@@

\subsection{Conditional Use Accuracy Equality}
Conditional use accuracy equality \cite{berk2021fairness} is similar to equalized odds, but instead of conditioning on the ground truth, it conditions on the prediction and calculated the probability of the ground truth. It can be seem as checking the prediction accuracy across groups, thus the name. It further requires the measure on the case of positive predictive values is less than that of the negative predictive values. Formally, it is:

\begin{align*}
    \abs{P[Y = 1 | S = 1, \hat{Y} = 1] & - P[Y = 1 | S \neq 1, \hat{Y} = 1]} && \leq \epsilon \\
    & \rotatebox{90}{>} && \\
    \abs{P[Y = 0 | S = 1, \hat{Y} = 0] & - P[Y = 0 | S \neq 1, \hat{Y} = 0]} && \leq \epsilon
\end{align*}

\subsection{Predictive Parity}
Predictive parity \cite{chouldechova2017fair} is a relaxation of conditional use accuracy equality by only considering the positive predictive value case. Like predictive equality, it was formulated by considering positive outcome being detrimental to the individual. Thus, only the positive predictive value case is taken. Formally, it is:

\begin{align*}
    \abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon
\end{align*}

\subsection{Equal Calibration}
Equal calibration \cite{chouldechova2017fair} is similar to equal opportunity, but instead of having a binary $\hat{Y}$, it is conditioned on the range of the predicted value $\hat{V}$. For example, this could be conditioned on the highest COMPAS score $\hat{V} = 10$. Calibration is a concept of having fair score function \cite{fraenkel2020fairness}. Formally, it is:

\begin{align*}
    \abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon
\end{align*}

\subsection{Positive Balance}
Positive balance \cite{kleinberg2016inherent} is similar to equal opportunity, but instead of taking the difference of probability of binary prediction $\hat{Y}$, it takes the difference of the expected value of the score $\hat{V}$. Formally, it is:

\begin{align*}
    \abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon
\end{align*}

\subsection{Negative Balance}
Negative balance \cite{kleinberg2016inherent} is like positive balance except it conditions on the case of $Y = 0$. Formally, it is:

\begin{align*}
    \abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon
\end{align*}

These fairness measures are compiled in Table~\ref{tab:measures}.

\begin{table*}[h]
    \centering
    % \resizebox{\textwidth}{!}{
    % \resizebox{\columnwidth}{!}{
    \begin{tabular}{|l|c|} % 'l' specifies left alignment for the first column
        \hline
        \textbf{Column 1} & \textbf{Column 2} \\
        \hline
Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
\hline
Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
\hline
Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
\hline
Overall Accuracy Eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
\hline
Mean Difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
\hline
\multirow{2}{*}{Equalized Odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\cline{2-2}
                                & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
\hline
Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
\hline
Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\hline
\multirow{2}{*}{Conditional Use Accuracy Equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
\cline{2-2}
                                 & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
\hline
Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
\hline
Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
\hline
Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
\hline
Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
\hline
    \end{tabular}
    % }
    \caption{Fairness measures.}
    \label{tab:measures}
\end{table*}

\section{Auditing Framework}
Per the review by Pessach and Shmueli \cite{pessach2022review}, we designed an auditing framework for calculating the various fairness measure. We offer two versions of fairness checkers: one for when the prediction positive results are readily available in csv input and one for when a model is provided. In most auditing cases the model version is preferred because csv result can be easily fabricated.

\subsection{Preliminaries}
Let $\db = \set{r_1, r_2, ...}$ be a database containing rows of data. Each row $r_i : \mathtt{str} \implies \mathtt{str}$ is a lookup table or dictionary. For example, $r_n(\text{``race''}) = \text{``Asian''}$ means $r_n$'s race is Asian.

A model $\model : (\mathtt{str} \implies \mathtt{str}) \implies \alpha$ takes a row and returns the model's prediction result of some type $\alpha$.

A privileged predicate $\priv : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if it belongs to the privileged group.

A positive predicate of rows $\pos : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if its prediction is positive. A positive predicate of model results $\pos : \alpha \implies \mathtt{bool}$ takes a model's result and determines if it is positive.

A ground truth predicate $\tru : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and gives the ground truth of the result.

A legitimate predicate $\leg : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, and similarly a score predicate $\sco : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, takes $n$ parameters and returns a row predicate.

\subsection{Definitions}
We abstracted the idea of privileged groups and positive prediction as predicates to maximizes flexibility. With the proper predicates, any model output can be used; even prose like response of generative models can be included given adequate predicates.

Given a datasets for auditing use, and given csv prediction results or the prediction model itself, we can calculate the fairness measures by:

\begin{align*}
\text{disparate\_impact} & (\epsilon, \model, \priv, \pos) \\
\text{demographic\_parity} & (\epsilon, \model, \priv, \pos) \\
\text{conditional\_statistical\_parity} & (\epsilon, \model, \priv, \pos, \leg, (args, ...)) \\
\text{overall\_accuracy\_eqaulity} & (\epsilon, \model, \priv, \pos, \tru) \\
\text{mean\_difference} & (\epsilon, \model, \priv, \pos) \\
\text{equalized\_odds} & (\epsilon, \model, \priv, \pos, \tru) \\
\text{equal\_opportunity} & (\epsilon, \model, \priv, \pos, \tru) \\
\text{predictive\_equality} & (\epsilon, \model, \priv, \pos, \tru) \\
\text{conditional\_use\_accuracy\_equality} & (\epsilon, \model, \priv, \pos, \tru) \\
\text{predictive\_parity} & (\epsilon, \model, \priv, \pos, \tru) \\
\text{equal\_calibration} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
\text{positive\_balance} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
\text{nagative\_balance} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
\end{align*}

where $\mathcal{M}$ is optional if the input contains csv prediction results.
